---
title: "Tracing Civil War Journeys"
format:
  html:
    toc: false
    code-fold: false
    page-layout: full
    theme: cosmo
    css: styles.css
    smooth-scroll: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
library(dplyr)
library(readr)
library(stringr)
library(tidyr)
library(tibble)
library(leaflet)
library(htmltools)
library(sf)
library(purrr)
library(tidygeocoder)
library(htmlwidgets)
```

## About this project

::: {.callout-note appearance="simple"}
This interactive map traces documented and approximate journey stops for soldiers and sailors in the dataset. It is designed as a public-facing research tool: readers can explore movement across enlistment points, transfers, receiving ships, discharge locations, and other recorded stops.

Please note that the lines connecting to each stop are not representative of the exact journey the individual might have taken, but rather a visualization of distance.
:::

This first version of the map emphasizes **documented locations** and **carefully standardized place names** rather than speculative routes. Lines connect each person's stops in order using `stop_order`.

### How to read the map

-   **Lines** trace each recorded journey in sequence.
-   **Point colors** distinguish exact, approximate, and uncertain locations.
-   **Line colors** distinguish Army and Navy records.
-   Clicking a point opens source-based details for that stop.

```{r}
#| label: load-data
stops_raw <- read_csv("data/soldier_data_mapping_ready.csv", show_col_types = FALSE)

stops <- stops_raw %>%
  mutate(
    place_standardized = na_if(str_squish(place_standardized), ""),
    place_label = na_if(str_squish(place_label), ""),
    stop_date = na_if(as.character(stop_date), ""),
    certainty = coalesce(certainty, "unknown"),
    service_branch = coalesce(service_branch, "Unknown")
  ) %>%
  filter(!is.na(place_standardized))
```

```{r}
#| label: geocode-cache
#| eval: true
geocode_cache_path <- "data/place_coordinates.csv"

if (file.exists(geocode_cache_path)) {
  geocode_cache <- read_csv(geocode_cache_path, show_col_types = FALSE)
} else {
  geocode_cache <- tibble(
    place_standardized = character(),
    lat = numeric(),
    lon = numeric()
  )
}

places_needed <- stops %>%
  distinct(place_standardized) %>%
  anti_join(geocode_cache, by = "place_standardized")

if (nrow(places_needed) > 0) {
  new_geocodes <- places_needed %>%
    geocode(
      address = place_standardized,
      method = "osm",
      lat = lat,
      long = lon,
      limit = 1
    ) %>%
    select(place_standardized, lat, lon)

  geocode_cache <- bind_rows(geocode_cache, new_geocodes) %>%
    distinct(place_standardized, .keep_all = TRUE)

  write_csv(geocode_cache, geocode_cache_path)
}

stops_geo <- stops %>%
  select(-lat, -lon) %>%
  left_join(geocode_cache, by = "place_standardized")
```

```{r}
#| label: prepare-map-data
stops_map <- stops_geo %>%
  filter(!is.na(lat), !is.na(lon)) %>%
  arrange(name, stop_order) %>%
  mutate(
    popup = pmap_chr(
      list(name, service_branch, stop_type, stop_date, place_label, place_standardized, certainty, notes_for_mapping),
      ~ as.character(tagList(
        tags$b(..1), tags$br(),
        tags$strong("Branch: "), ..2, tags$br(),
        tags$strong("Stop type: "), ..3, tags$br(),
        tags$strong("Date: "), ifelse(is.na(..4), "Unknown", ..4), tags$br(),
        tags$strong("Original place: "), ifelse(is.na(..5), "Unknown", ..5), tags$br(),
        tags$strong("Mapped place: "), ifelse(is.na(..6), "Unknown", ..6), tags$br(),
        tags$strong("Certainty: "), ..7, tags$br(),
        tags$strong("Notes: "), ifelse(is.na(..8), "", ..8)
      ))
    ),
    stop_group = paste0("Stops: ", name)         # CHANGED: group name for each soldier's stops
  )

segments <- stops_map %>%
  group_by(name) %>%
  arrange(stop_order, .by_group = TRUE) %>%
  mutate(
    next_lat = lead(lat),
    next_lon = lead(lon),
    next_place = lead(place_standardized),
    next_stop_order = lead(stop_order)
  ) %>%
  ungroup() %>%
  filter(!is.na(next_lat), !is.na(next_lon))

segments_sf <- pmap_dfr(
  list(
    segments$lon, segments$lat, segments$next_lon, segments$next_lat,
    segments$name, segments$service_branch, segments$place_standardized, segments$next_place
  ),
  function(lon, lat, next_lon, next_lat, name, service_branch, place_standardized, next_place) {
    tibble(
      name = name,
      service_branch = service_branch,
      place_standardized = place_standardized,
      next_place = next_place,
      line_group = paste0("Journey: ", name),   # CHANGED: group name for each soldier's journey line
      geometry = st_sfc(
        st_linestring(matrix(c(lon, lat, next_lon, next_lat), ncol = 2, byrow = TRUE)),
        crs = 4326
      )
    )
  }
) %>%
  st_as_sf()

branch_pal <- colorFactor(
  palette = c("Army" = "#8c510a", "Navy" = "#01665e", "Unknown" = "#5e3c99"),
  domain = unique(stops_map$service_branch)
)

certainty_pal <- colorFactor(
  palette = c("exact" = "#1b9e77", "approximate" = "#d95f02", "unknown" = "#7570b3"),
  domain = unique(stops_map$certainty)
)

```

## Explore the map

Use the interactive controls to zoom, pan, and compare Army and Navy journeys. You can turn stops and journey lines on or off in the layer panel.

```{r}
#| label: leaflet-map

# Create slightly offset coordinates only where points overlap
stops_map_jittered <- stops_map %>%
  group_by(lon, lat) %>%
  mutate(
    n_at_location = n(),
    row_id_at_location = row_number(),
    angle = ifelse(n_at_location == 1, 0, 2 * pi * (row_id_at_location - 1) / n_at_location),
    offset = ifelse(n_at_location == 1, 0, 0.18),
    lon_jitter = lon + offset * cos(angle),
    lat_jitter = lat + offset * sin(angle)
  ) %>%
  ungroup()

# CHANGED: save all soldier names for the dropdown
soldier_names <- sort(unique(stops_map_jittered$name))

# CHANGED: start leaflet map as an object so we can attach dropdown JS after building it
m <- leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron)

# CHANGED: add journey lines soldier-by-soldier so each one can be selected in the dropdown
for (soldier in soldier_names) {
  m <- m %>%
    addPolylines(
      data = segments_sf %>% filter(name == soldier),
      color = ~branch_pal(service_branch),
      weight = 2.5,
      opacity = 0.7,
      group = paste0("Journey: ", soldier),
      popup = ~paste0(
        "<b>", name, "</b><br>",
        "From: ", place_standardized, "<br>",
        "To: ", next_place
      )
    )
}

# CHANGED: add stop points soldier-by-soldier so each one can be selected in the dropdown
for (soldier in soldier_names) {
  m <- m %>%
    addCircleMarkers(
      data = stops_map_jittered %>% filter(name == soldier),
      lng = ~lon_jitter,
      lat = ~lat_jitter,
      radius = 6,
      stroke = TRUE,
      weight = 1,
      color = ~certainty_pal(certainty),
      fillColor = ~certainty_pal(certainty),
      fillOpacity = 0.9,
      popup = ~popup,
      group = paste0("Stops: ", soldier),
      label = ~paste(name, "-", stop_type)
    )
}

# keep legends and regular layer control
m <- m %>%
  addLegend(
    position = "bottomright",
    pal = branch_pal,
    values = stops_map$service_branch,
    title = "Service Branch",
    opacity = 1
  ) %>%
  addLegend(
    position = "bottomleft",
    pal = certainty_pal,
    values = stops_map$certainty,
    title = "Location Certainty",
    opacity = 1
  ) %>%
  addLayersControl(
    overlayGroups = c("Show all journeys", "Show all stops"),
    options = layersControlOptions(collapsed = FALSE)
  )

# CHANGED: add dropdown menu with JavaScript
m <- htmlwidgets::onRender(
  m,
  sprintf("
function(el, x) {
  var map = this;

  // --- CHANGED: build dropdown control ---
  var control = L.control({position: 'topright'});
  control.onAdd = function(map) {
    var div = L.DomUtil.create('div', 'info leaflet-bar');
    div.style.background = 'white';
    div.style.padding = '6px';
    div.style.borderRadius = '4px';
    div.style.boxShadow = '0 1px 5px rgba(0,0,0,0.4)';

    div.innerHTML = `
      <label for='soldier-select' style='font-weight:600; display:block; margin-bottom:4px;'>
        Select a soldier
      </label>
      <select id='soldier-select' style='width:220px;'>
        <option value='ALL'>All soldiers</option>
        %s
      </select>
    `;

    L.DomEvent.disableClickPropagation(div);
    return div;
  };
  control.addTo(map);

  function setVisibility(selected) {
    var byGroup = map.layerManager._byGroup;

    Object.keys(byGroup).forEach(function(groupName) {
      if (groupName.startsWith('Journey: ') || groupName.startsWith('Stops: ')) {
        var show = (selected === 'ALL') ||
                   (groupName === 'Journey: ' + selected) ||
                   (groupName === 'Stops: ' + selected);

        Object.keys(byGroup[groupName]).forEach(function(id) {
          var layer = byGroup[groupName][id];
          if (show) {
            map.addLayer(layer);
          } else {
            map.removeLayer(layer);
          }
        });
      }
    });
  }

  document.getElementById('soldier-select').addEventListener('change', function(e) {
    setVisibility(e.target.value);
  });
}
",
  paste0("<option value=\\'", soldier_names, "\\'>", soldier_names, "</option>", collapse = "")
))

m

```
