
---
title: "Tracing Civil War Journeys"
format:
  html:
    toc: false
    code-fold: false
    page-layout: full
    theme: cosmo
    css: styles.css
    smooth-scroll: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
library(dplyr)
library(readr)
library(stringr)
library(tidyr)
library(tibble)
library(leaflet)
library(htmltools)
library(sf)
library(purrr)
library(tidygeocoder)
```

## About this project

::: {.callout-note appearance="simple"}
This interactive map traces documented and approximate journey stops for soldiers and sailors in the dataset. It is designed as a public-facing research tool: readers can explore movement across enlistment points, transfers, receiving ships, discharge locations, and other recorded stops.
:::

This first version of the map emphasizes **documented locations** and **carefully standardized place names** rather than speculative routes. Lines connect each person's stops in order using `stop_order`.

### How to read the map

-   **Lines** trace each recorded journey in sequence.
-   **Point colors** distinguish exact, approximate, and uncertain locations.
-   **Line colors** distinguish Army and Navy records.
-   Clicking a point opens source-based details for that stop.

```{r}
#| label: load-data
stops_raw <- read_csv("data/journey_stops_mapping_ready.csv", show_col_types = FALSE)

stops <- stops_raw %>%
  mutate(
    place_standardized = na_if(str_squish(place_standardized), ""),
    place_label = na_if(str_squish(place_label), ""),
    stop_date = na_if(as.character(stop_date), ""),
    certainty = coalesce(certainty, "unknown"),
    service_branch = coalesce(service_branch, "Unknown")
  ) %>%
  filter(!is.na(place_standardized))
```

## Geocode places once and cache them

This chunk geocodes only places that do not already have latitude and longitude. It saves a reusable lookup table so you do not have to geocode everything every time you render.

```{r}
#| label: geocode-cache
#| eval: true
geocode_cache_path <- "data/geocoded_places.csv"

if (file.exists(geocode_cache_path)) {
  geocode_cache <- read_csv(geocode_cache_path, show_col_types = FALSE)
} else {
  geocode_cache <- tibble(
    place_standardized = character(),
    lat = numeric(),
    lon = numeric()
  )
}

places_needed <- stops %>%
  distinct(place_standardized) %>%
  anti_join(geocode_cache, by = "place_standardized")

if (nrow(places_needed) > 0) {
  new_geocodes <- places_needed %>%
    geocode(
      address = place_standardized,
      method = "osm",
      lat = lat,
      long = lon,
      limit = 1
    ) %>%
    select(place_standardized, lat, lon)

  geocode_cache <- bind_rows(geocode_cache, new_geocodes) %>%
    distinct(place_standardized, .keep_all = TRUE)

  write_csv(geocode_cache, geocode_cache_path)
}

stops_geo <- stops %>%
  select(-lat, -lon) %>%
  left_join(geocode_cache, by = "place_standardized")
```

```{r}
#| label: prepare-map-data
stops_map <- stops_geo %>%
  filter(!is.na(lat), !is.na(lon)) %>%
  arrange(name, stop_order) %>%
  mutate(
    popup = pmap_chr(
      list(name, service_branch, stop_type, stop_date, place_label, place_standardized, certainty, notes_for_mapping),
      ~ as.character(tagList(
        tags$b(..1), tags$br(),
        tags$strong("Branch: "), ..2, tags$br(),
        tags$strong("Stop type: "), ..3, tags$br(),
        tags$strong("Date: "), ifelse(is.na(..4), "Unknown", ..4), tags$br(),
        tags$strong("Original place: "), ifelse(is.na(..5), "Unknown", ..5), tags$br(),
        tags$strong("Mapped place: "), ifelse(is.na(..6), "Unknown", ..6), tags$br(),
        tags$strong("Certainty: "), ..7, tags$br(),
        tags$strong("Notes: "), ifelse(is.na(..8), "", ..8)
      ))
    )
  )

segments <- stops_map %>%
  group_by(name) %>%
  arrange(stop_order, .by_group = TRUE) %>%
  mutate(
    next_lat = lead(lat),
    next_lon = lead(lon),
    next_place = lead(place_standardized),
    next_stop_order = lead(stop_order)
  ) %>%
  ungroup() %>%
  filter(!is.na(next_lat), !is.na(next_lon))

segments_sf <- pmap_dfr(
  list(segments$lon, segments$lat, segments$next_lon, segments$next_lat,
       segments$name, segments$service_branch, segments$place_standardized, segments$next_place),
  function(lon, lat, next_lon, next_lat, name, service_branch, place_standardized, next_place) {
    tibble(
      name = name,
      service_branch = service_branch,
      place_standardized = place_standardized,
      next_place = next_place,
      geometry = st_sfc(
        st_linestring(matrix(c(lon, lat, next_lon, next_lat), ncol = 2, byrow = TRUE)),
        crs = 4326
      )
    )
  }
) %>%
  st_as_sf()

branch_pal <- colorFactor(
  palette = c("Army" = "#8c510a", "Navy" = "#01665e", "Unknown" = "#5e3c99"),
  domain = unique(stops_map$service_branch)
)

certainty_pal <- colorFactor(
  palette = c("exact" = "#1b9e77", "approximate" = "#d95f02", "unknown" = "#7570b3"),
  domain = unique(stops_map$certainty)
)
```

## Explore the map

Use the interactive controls to zoom, pan, and compare Army and Navy journeys. You can turn stops and journey lines on or off in the layer panel.

```{r}
#| label: leaflet-map

# Create slightly offset coordinates only where points overlap
stops_map_jittered <- stops_map %>%
  group_by(lon, lat) %>%
  mutate(
    n_at_location = n(),
    row_id_at_location = row_number(),
    angle = ifelse(n_at_location == 1, 0, 2 * pi * (row_id_at_location - 1) / n_at_location),
    offset = ifelse(n_at_location == 1, 0, 0.18),
    lon_jitter = lon + offset * cos(angle),
    lat_jitter = lat + offset * sin(angle)
  ) %>%
  ungroup()

leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolylines(
    data = segments_sf,
    color = ~branch_pal(service_branch),
    weight = 2.5,
    opacity = 0.7,
    group = "Journeys",
    popup = ~paste0(
      "<b>", name, "</b><br>",
      "From: ", place_standardized, "<br>",
      "To: ", next_place
    )
  ) %>%
  addCircleMarkers(
    data = stops_map_jittered,
    lng = ~lon_jitter,
    lat = ~lat_jitter,
    radius = 6,
    stroke = TRUE,
    weight = 1,
    color = ~certainty_pal(certainty),
    fillColor = ~certainty_pal(certainty),
    fillOpacity = 0.9,
    popup = ~popup,
    group = "Stops",
    label = ~paste(name, "-", stop_type)
  ) %>%
  addLegend(
    position = "bottomright",
    pal = branch_pal,
    values = stops_map$service_branch,
    title = "Service Branch",
    opacity = 1
  ) %>%
  addLegend(
    position = "bottomleft",
    pal = certainty_pal,
    values = stops_map$certainty,
    title = "Location Certainty",
    opacity = 1
  ) %>%
  addLayersControl(
    overlayGroups = c("Journeys", "Stops"),
    options = layersControlOptions(collapsed = FALSE)
  )
```
